C Code Conversion to One Counter Automata for Reachability Analysis

### 2020-03-29 Documentation/paper.tex
updated paper

\usepackage{biblatex}
\addbibresource{bibliography.bib}

@@

\title{C Code Conversion to Counter Automata for Reachability Analysis}
\author{Lars Van Roy\\
\textit{dept. of Mathematics and Computer Science} \\
\textit{University of Antwerp}\\
lars.vanroy@student.uantwerpen.be}
\date{\today}

@@

\section{Introduction}
A well known problem in writing code is the reachability of specific lines in said written code. It is often impossible to properly determine whether or not a line of code guarded by a conditional statement is going to be reachable in any possible execution of the code\cite{713323, 10.1145/1292316.1292319}. Unreachable code is never desired, but is an ever more pressing issue for energy restrained environments such as embedded computer systems \cite{embeddedefficiency}. In these systems, we want our execution to be as efficient as possible, and therefore avoid conditions that can never be satisfied or lines that can never be reached at any cost. In order to upgrade the efficiency of code that is affected by time constraints, or code in general, a proper reachability analysis would be of great use.

In order to analyse code reachability, we will want to make use of one counter automata. The one counter automata are especially interesting as code is easily convertible to counter automata and it has been proven \cite{succintandparametriconecounter, danialandjoel} that reachability is decidable for this type of automata. As will be shown in the continuation of this paper, adding the restriction of only allowing a single parametric clock has a big effect on the types of code we can analyse, but is unavoidable, since there is no proof for two counter machines. Reachability problems for machines with three or more counters have been proven to be undecidable.

@@

We will only consider a subset of c programs, which contain at most one counter where a counter is a variable that is used inside a conditional statement. The program will be able to handle c code with zero counters, but this is considered trivial and will be disregarded for the remainder of this paper. Furthermore, the operations on the counter will be restricted, the counter can only be updated via assignments with parameters or constants, and the counter can only be compared using equality, inequality, strict less than, strict greater than, less than or equal, greater than or equal, and only to parameters or constants. Furthermore, the function parameters are not allowed to be altered throughout the duration of the code.

@@

This proves that there is a definite need for additional research on analysis in code. Reachability analysis by using counter automata would not only allow for an expansion of the current reachability analysis as is, but might also allow further expansions based on the results of the counter automaton expansion.

@@

\subsection{General automata}
The automata used in this research are a specific subset of the generally defined automata. The general definition for an automaton A is given by the tuple (Q, $\Sigma$, $\delta$, $q_0$, F) where
\begin{itemize}
	\item Q is the set of states
	\item $\Sigma$ is a finite set of symbols, called the alphabet of the automaton
	\item $\delta$ is the transition function, $\delta: Q \times \Sigma \rightarrow Q$
	\item $q_0$ is the initial state, $q_0 \in Q$
	\item F is a set of states of Q defining the final states, $F \subseteq Q$
\end{itemize}

\subsection{Counter automata}

@@

The counter valuation $c \in \mathbb{N}$ is the value that will be analysed, updated during the evaluation of the automaton.

@@

Finally, we will consider the following set of allowed operations where $P$ is the set of parameters, the $\epsilon$ symbol will be used to represent an empty label.
\[
 op = \{+c, -c, +p, -p, \leq c, =c, \geq c, \leq p, =p, \geq p: c \in \mathbb{N}, p \in P\} \cup \{\epsilon\}
\]

The automata we will consider will be one-counter automata. The set $Q$ will contain a state for each line of code. The initial state $q_0$ will be the state corresponding with the first line of code. For now the set of final states $F$ will be empty, but the intended interpretation of a final state will be the line of which the reachability will be checked. The transition relation will consist of elements of the form $(q_i, l, q_{i+1})$ where $q_i, q_{i+1} \in Q$ and $l \in op$.

@@

\subsubsection{Context-sensitive reduction example: addition folding}
As described earlier, we do allow additions and subtractions of the form $-x$ and $+x$, where $x \in P$ or $x \in \mathbb{N}$ with $P$ the set of parameters. Given that the formation described in Figure \ref{fig:unfolded_addition} occurs in the automaton, we would technically not be allowed to accept this, as this is not directly in the specified format.

@@

\subsubsection{Assignments}
Other than functions, we will also need to know when assignments to counters occur, so that we can add the proper transition labels. We only need to check whether or not we are assigning to a counter, if so, we read the operation used, and the variable that gets assigned. Valid counter assignments need to be of the form \textit{c op x} where $c$ is the counter, $op \in $ \textit{\{=, +=, -=\}} and $x \in P$ or $x \in \mathbb{N}$.

@@

\begin{figure}[h]
	\centering
	\includegraphics[width=\linewidth]{counter_switch}
	\caption{Example of a switch statement.}
	\label{fig:counter_switch}
\end{figure}
\subsubsection{If statements}
\label{sec:if}
Another important construct are the if statements. These statements will, just as the iteration statements did, require additional nodes to be able to be modelled. If statements require a total of 4 nodes to be modelled.

@@

\subsubsection{less than and greater than}
\label{sec:strict}
The final addition to the set of supported statements, are the less than and greater than expressions. These are needed to model the opposing conditions to greater than or equal and less than or equal respectively.

@@

@@

\subsubsection{Final overview}
To give a in depth overview of how the conversion works, we will transform the following code into a counter automaton. This code consists of multiple if statements, as well as a loop, it makes use of a single counter value, being \textit{counter}, and two parameters. The loop and the final if both use inequality expressions, which are representable in a counter automaton, as was described earlier. The first two ifs make use of relational expressions with the $<$ and $>$ operators, which are also representable within a counter automaton. The counter is only modified using additions or subtractions with parameters and none of these parameters will ever be modified, which also conforms to the requirements. Finally, this function has only integer parameters and a bool return type, which makes us conclude that this entire function is representable in a counter automaton.

@@

\section{Tests}
The referenced tool has been thoroughly tested using three groups of tests, which are all white box tests. The tests will check the outputted dot and the trace output of the program.

@@

The third group will test the full tool. It has a few happy day scenarios, where all kinds of statements get tested in multiple configurations. As well as a lot of tests where invalid statements occur, where we test whether or not the program is capable of detecting all issues in the code.

\section{Conclusion}
In this paper we gave a brief description of a program capable of converting a small subset of the c language to counter automata. The resulting automata can be used in reachability analysis within the code.

@@

The usability of this program without any further additions is very limited. The restrictions enforced on the c code make it fairly useless in more complex situations, where reachability analysis would be far more desired however, it is very effective in case the program is applicable. Big conditional statements or a large amount of conditional statements can make the reachability of certain sub segments hard, and with the addition of a reachability checker, they could definitely lead to improvements in coding.

>>>>

\usepackage[backend=bibtex]{biblatex}
\addbibresource{bibliography}

@@

\title{C Code Conversion to One Counter Automata for Reachability Analysis}
\author{Lars Van Roy\\
\textit{dept. of Mathematics and Computer Science} \\
\textit{University of Antwerp}\\
lars.vanroy@student.uantwerpen.be}
\date{\today}

@@

\section{Introduction}
A well known problem in writing code is the reachability of specific lines in said written code. It is often impossible to properly determine whether or not a line of code guarded by a conditional statement is going to be reachable in any possible execution of the code \cite{713323, 10.1145/1292316.1292319}. Unreachable code is never desired, but is an ever more pressing issue for energy restrained environments such as embedded computer systems \cite{embeddedefficiency}. In these systems, we want our execution to be as efficient as possible, and therefore avoid conditions that can never be satisfied or lines that can never be reached at any cost. In order to upgrade the efficiency of code that is affected by time constraints, or code in general, a proper reachability analysis is of great use.

To analyze reachability in code artificats, we propose a translation from code to one counter automata. The one counter automata are especially interesting as code is easily convertible to counter automata and it has been proven that reachability is decidable for this type of automata \cite{succintandparametriconecounter, danialandjoel}. As will be shown in the continuation of this paper, adding the restriction of only allowing a single parametric clock has a big effect on the types of code we can analyse, but is unavoidable, since reachability for two counter machines has been proven to be np hard and reachability problems for machines with three or more counters have been proven to be undecidable \cite{FEARNLEY201526}.

@@

We will only consider a subset of c programs, which contain at most one counter. Here a counter is a variable that is used inside a conditional statement. The program will be able to handle c code with zero counters, but this is considered trivial and will be disregarded for the remainder of this paper. Furthermore, the operations on the counter will be restricted, the counter can only be updated via assignments with parameters or constants, and the counter can only be compared using equality, inequality, strict less than, strict greater than, less than or equal, greater than or equal, and only to parameters or constants. Furthermore, the function parameters are not allowed to be altered throughout the duration of the code.

@@

This shows that there is a definite need for additional research on analysis in code. Reachability analysis by using counter automata would allow to detect more cases of dead code than the original dead code analysis techniques could offer.

@@

\subsection{General automata}
The automata used in this research are a specific subset of the generally defined automata. The general definition for an automaton A is given by the tuple (Q, $\Sigma$, $\delta$, $q_0$, F) where
\begin{itemize}
	\item Q is the set of states
	\item $\Sigma$ is a finite set of symbols, called the alphabet of the automaton
	\item $\delta: Q \times \Sigma \rightarrow Q$ is the transition function
	\item $q_0 \in Q$ is the initial state
	\item $F \subseteq Q$ is a set of states of Q defining the final states, 
\end{itemize}

\subsection{Parametric counter automata}

@@

The counter valuation $c \in \mathbb{N}$ is the value that will be analysed and updated during the evaluation of the automaton.

@@

Parametric counter automata, as compared to regular counter automata, allow counters to be modified by parameters as well as constants. As the counters need to be integers, we will only allow integer type paramters. We will consider the following set of allowed operations where $P$ is the set of parameters, the $\epsilon$ symbol will be used to represent an empty label.
\[
 op = \{+c, -c, +p, -p, \leq c, =c, \geq c, \leq p, =p, \geq p: c \in \mathbb{N}, p \in P\} \cup \{\epsilon\}
\]

The automata we will consider will be one-counter parametric automata. The set $Q$ will contain a state for each line of code. The initial state $q_0$ will be the state corresponding with the first line of code. For now the set of final states $F$ will be empty, but the intended interpretation of a final state will be the line of which the reachability will be checked. The transition relation will consist of elements of the form $(q_i, l, q_{i+1})$ where $q_i, q_{i+1} \in Q$ and $l \in op$.

@@

\subsubsection{Context-sensitive reduction example: addition folding}
As described earlier, we do allow additions and subtractions of the form $-x, +x, -n$ and $+n$, where $x \in P$ or $n \in \mathbb{N}$ with $P$ the set of parameters. Given that the formation, described in Figure \ref{fig:unfolded_addition} occurs in the automaton, we would technically not be allowed to accept this, as this is not directly in the specified format.

@@

\subsubsection{Assignments}
Other than functions, we will also need to know when assignments to counters occur, so that we can add the proper transition labels. We only need to check whether or not we are assigning to a counter, if so, we read the operation used, and the variable that gets assigned. Valid counter assignments need to be of the form \textit{c op x} where $c$ is the counter, $op \in $ \textit{\{=, +=, -=\}} and $x \in P \lor x \in \mathbb{N}$.

@@

@@

\subsubsection{Less than and greater than}
\label{sec:strict}
The final addition to the set of supported statements, are the less than and greater than expressions. These are needed to model the opposing conditions to greater than or equal and less than or equal respectively.

@@

\subsubsection{If statements}
\label{sec:if}
Another important construct are the if statements. These statements will, just as the iteration statements did, require additional nodes to be able to be modelled. If statements require a total of 4 nodes to be modelled.

\begin{figure}[h]
	\centering
	\includegraphics[width=\linewidth]{counter_switch}
	\caption{Example of a switch statement.}
	\label{fig:counter_switch}
\end{figure}

@@

\subsubsection{Final overview}
To give a in depth overview of how the conversion works, we will transform the following code into a counter automaton. This code consists of multiple if statements, as well as a loop, it makes use of a single counter value, being \textit{counter}, and two parameters. The loop and the final \textit{if} both use inequality expressions, which are representable in a counter automaton, as was described earlier. The first two \textit{ifs} make use of relational expressions with the $<$ and $>$ operators, which are also representable within a counter automaton. The counter is only  using additions or subtractions with parameters and none of these parameters will ever be modified, which also conforms to the requirements. Finally, this function has only integer parameters and a bool return type, which makes us conclude that this entire function is representable in a counter automaton.

@@

\section{Tool Validation}
To determine the corectness of our tool, we created a test suite consisting out of three groups of tests, which are all white box tests. The tests will check the outputted dot and the trace output of the program, against traces and dot files which have been manually generated and verified.

@@

The third group will test the full tool. It will validate the dot file resulting from the cleaner, as well as the dot file(s) resulting from the counter automaton generator. It will also evaluate all output regarding the symbol table, and the potential errors of the code validator. There are happy day tests where each conditional statement gets tested separately, there are tests where conditional statements get nested and there are tests to see whether we can still evaluate conditional statements with constant conditions. Furthermore, there are failure tests that test whether or not we can properly detect incorrect counter types, incorrect parameter types, incorrect number of counters, incorrect conditional operations, incorrect counter modifications and properly detect counters defined in global scopes. There are also tests for edge case scenarios, such as returns in every single branch, the cleaner will not be able to detect these, and therefore the generator should be able to stop generating all nodes, once there are no branch that can carry on.

The test coverage was validated using Codecov, which gave a total line coverage of 93\% at the end of development. This coverage was used as a metric to determine whether or not there were sufficient tests at each stage of development. The remaining 7\% are trivial lines that did not need any testing, such as \textit{\_\_str\_\_} operators.

\section{Conclusion}
In this paper we propose an algorithm for converting a small subset of the c language to counter automata, and discuss its implementation. The resulting automata can be used in reachability analysis within the code.

@@

This proposed algorithm can be used to improve existing static code analysis and will furthermore be usable to solve the halting problem for a subset of c programs. It is however only usable in a limited amount of cases, in more complex situations, where reachability analysis would be far more desired, it will most likely not be usable as chances of violated conditions will be high. In the situations where it is applicable, big conditional statements or a large amount of conditional statements can make determining the reachability of certain sub segments hard, and with the addition of this algorithm as a reachability checker, it could definitely lead to improvements in coding.
